\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{multirow}
\geometry{margin=2.5cm}

\definecolor{codebg}{RGB}{245,245,245}
\definecolor{crysblue}{RGB}{30,80,180}
\lstset{
  backgroundcolor=\color{codebg},
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  rulecolor=\color{gray},
  keywordstyle=\color{crysblue}\bfseries,
  commentstyle=\color{gray}\itshape,
}

\newtheorem{definition}{Definition}
\newtheorem{property}{Property}

\title{\textbf{Qomni Engine v8: Adaptive Early-Exit Cognitive Cascade (AECC)\\
for Minimum-Cost Domain Inference}\\
\large Preprint --- April 2026}

\author{
  Percy Rojas Masgo\textsuperscript{1} \quad Qomni AI Lab\textsuperscript{2}\\[0.3em]
  \textsuperscript{1}Condesi Per\'u \quad
  \textsuperscript{2}Qomni AI Lab\\
  \href{mailto:percy@condesi.pe}{percy@condesi.pe}
}
\date{April 2026}

\begin{document}
\maketitle

% ─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
We present \textbf{Qomni Engine v8} and the \textbf{Adaptive Early-Exit
Cognitive Cascade (AECC)}, an inference architecture that minimizes unnecessary
neural generation by cascading from zero-compute reflexes and deterministic
programs to indexed retrieval, and only when required, full LLM inference.

The central contribution is a \textbf{minimum-cost strategy selection} policy:
given a query $q$, Qomni selects the strategy $s^*(q)$ that minimizes
computational cost $C(s,q)$ subject to a quality constraint $Q(s,q) \geq \tau$.
In v8, this policy is approximated by a continuous confidence score
combined with an adaptive learning loop (Phase~0.59) and is designed to
extend to per-domain exponential-decay calibration (Phase~0.60).

Live benchmarks on commodity hardware (AMD EPYC 12-core, 48~GB RAM) across
12 domain-specific queries show \textbf{100\% LLM bypass rate}, \textbf{193~ms
median latency} vs.\ 8,000--45,000~ms for LLM inference, and
\textbf{100\% exact answers} on calculation queries.

We also present \textbf{CRYS-L v2}, an open domain-specific language for
deterministic engineering calculations compiled to WASM, released under
MIT license at \url{https://github.com/condesi/crysl-lang}.
\end{abstract}

\tableofcontents
\newpage

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

\paragraph{The LLM-first cost problem.}
The dominant paradigm for AI applications routes every query through a large
neural network regardless of the query's actual complexity. For
applications with structured domains --- engineering calculations, legal
databases, accounting rules --- this is orders of magnitude more costly than
necessary. A query such as ``how many HP for a 750~GPM fire pump at 100~PSI?''
has a \emph{closed-form, standard-referenced answer} computable in microseconds.
Routing it through a billion-parameter LLM consumes $\sim$8--45~seconds and
produces a probabilistic approximation with 15--25\% numerical error.

\paragraph{Execution-first inference.}
The right framing is not ``which LLM should I use?'' but
``what is the minimum computation required to resolve this query at acceptable quality?''
We call this \emph{execution-first inference}: deterministic computation,
pattern matching, and indexed retrieval are preferred over neural generation
unless confidence falls below a learned threshold.

\paragraph{Qomni Engine and AECC.}
\textbf{Qomni Engine} implements execution-first inference through the
\textbf{Adaptive Early-Exit Cognitive Cascade (AECC)}: a seven-strategy
pipeline where each strategy exits immediately upon sufficient confidence,
avoiding all downstream computation. Over eight major versions the system
evolved from a single-strategy distillation experiment to a 207-feature
cognitive orchestrator with a continuous learning loop.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{Formal problem statement} (\S\ref{sec:problem}): minimum-cost
    strategy selection as a constrained optimization, with confidence scoring
    as an approximation.
  \item \textbf{Seven-strategy cascade} (\S\ref{sec:cascade}): Reflex,
    CRYS-L JIT, Oracle-Direct, Crystal-Direct, Pipeline-Direct, Crystal-Ctx, Free LLM.
  \item \textbf{Phase 0.55--0.59 planner} (\S\ref{sec:planner}): binary
    $\to$ continuous confidence $\to$ adaptive learning loop.
  \item \textbf{Phase 0.60 design} (\S\ref{sec:p060}): per-domain EMA decay,
    Wilson-score calibration, multi-path preference learning.
  \item \textbf{CRYS-L v2} (\S\ref{sec:crysl}): open DSL for engineering
    calculations, MIT licensed, WASM-compiled.
  \item \textbf{Evaluation} (\S\ref{sec:eval}): four-metric evaluation framework
    and live benchmark data.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Problem Formulation}
\label{sec:problem}

\subsection{Minimum-Cost Strategy Selection}

Let $S = \{s_1, \ldots, s_k\}$ be the set of available inference strategies,
$q$ a query, $C(s,q) \geq 0$ the computational cost of executing strategy $s$
on $q$, and $Q(s,q) \in [0,1]$ a quality score for the result.
Let $\tau \in (0,1]$ be an acceptable quality threshold.

\begin{definition}[Optimal strategy]
\begin{equation}
  s^*(q) = \argmin_{s \in S} C(s, q)
  \quad \text{subject to} \quad Q(s, q) \geq \tau
  \label{eq:opt}
\end{equation}
\end{definition}

In practice, $Q(s,q)$ cannot be evaluated before executing $s$.
Qomni approximates equation~\eqref{eq:opt} with a learned confidence
function $\hat{q}(s, q)$ that predicts quality from observable query features
(keyword overlap, numeric context, domain match).

The cascade architecture instantiates a \emph{greedy exit policy}: strategies
are evaluated in increasing cost order $C(s_1,q) \leq C(s_2,q) \leq \ldots$,
and the engine exits at the first $s_i$ for which
$\hat{q}(s_i, q) \geq \theta_i$, where $\theta_i$ is a per-strategy threshold.

\subsection{Cost and Quality Models}

\paragraph{Cost.} For each strategy $s_i$, $C(s_i, q)$ is primarily wall-clock
latency. In decreasing efficiency order:
\[
  C(\text{Reflex}) < C(\text{CRYS-L}) < C(\text{Oracle/Crystal}) < C(\text{Pipeline}) < C(\text{Ctx-LLM}) \ll C(\text{Free-LLM})
\]

\paragraph{Quality.} We define quality along four dimensions (Section~\ref{sec:metrics}):
exactness, standard validity, reproducibility, and information coverage.
Deterministic strategies (Reflex, CRYS-L) achieve perfect exactness and
reproducibility at the expense of coverage; LLM achieves full coverage at
the expense of exactness.

\subsection{What Makes Qomni a Cognitive OS}

Three operational properties distinguish AECC from a simple router
and justify the term ``Adaptive Cognitive OS'':

\begin{property}[Multi-strategy execution]
At inference time, more than one strategy may execute concurrently.
Specifically, strategies 3--5 (Oracle, Crystal, Pipeline) execute as
\texttt{tokio::join!(oracle\_fut, crystal\_fut)}: whichever returns a valid
result first wins; both results are merged if both succeed.
\end{property}

\begin{property}[Adaptive learning loop]
After every query, the outcome is evaluated (validate\_result), recorded
(AtomicU64 counters), and used to adjust per-route thresholds.
The system improves its routing decisions without retraining.
\end{property}

\begin{property}[Persistent decision policy]
Learned thresholds survive restarts via atomic persistence
(\texttt{planner\_stats.json}). The system does not reset to factory defaults;
it continues from its accumulated routing experience.
\end{property}

These three properties together define the \textbf{Adaptive Early-Exit
Cognitive Cascade (AECC)}: a system that exits at the cheapest sufficient
strategy, learns from routing outcomes, and maintains persistent policy state
across restarts. The term \emph{Cognitive OS} reflects that the engine manages
inference resources across multiple modalities (computation, retrieval, neural),
analogously to how an operating system manages hardware resources.

% ─────────────────────────────────────────────────────────────────────────────
\section{Seven-Strategy Cascade}
\label{sec:cascade}

\begin{table}[h]
\centering
\caption{Qomni v8 strategy cascade}
\label{tab:cascade}
\begin{tabular}{clllrr}
\toprule
\# & Strategy & Mechanism & LLM & Latency & vs.\ Free-LLM \\
\midrule
1 & \textbf{Reflex} & HDC pattern + cached response & Never & $<5$~ms & $\mathbf{9{,}000\times}$ \\
2 & \textbf{CRYS-L JIT} & WASM engineering plan & Never & 50--200~ms & $\mathbf{225\times}$ \\
3 & \textbf{Oracle-Direct} & Formula index retrieval & Never & 100--400~ms & $\mathbf{112\times}$ \\
4 & \textbf{Crystal-Direct} & Knowledge KB retrieval & Never & 100--400~ms & $\mathbf{112\times}$ \\
5 & \textbf{Pipeline} & Oracle + Crystal parallel & Never & 100--500~ms & $\mathbf{90\times}$ \\
6 & \textbf{Crystal-Ctx} & Crystal context $\to$ LLM & Yes & 500~ms+ & $2\text{--}10\times$ \\
7 & \textbf{Free LLM} & Full neural generation & Yes & 8--45~s & baseline \\
\bottomrule
\end{tabular}
\end{table}

Strategies 1--5 never invoke the LLM. The cascade exits as soon as a strategy
returns a result that passes \texttt{validate\_result}. Strategy 7 is the
fallback of last resort.

\paragraph{CRYS-L JIT speed.}
The key to Strategy 2's speed is that WASM modules are compiled \emph{once}
and cached; subsequent executions are $<1$~ms of native arithmetic:

\begin{verbatim}
Query: "cuantos hp bomba nfpa20 750 gpm 100 psi"
  1. Plan match: "gpm", "psi", "bomba" -> plan_pump_sizing   (2ms)
  2. WASM execute: HP = (750x0.063 x 100x0.703)/(0.7x76)   (0.18ms)
  3. Format + cite: "18.04 HP — NFPA 20:2022 §4.26"        (0.1ms)
  Total end-to-end (HTTP+JSON+network):                     148ms
\end{verbatim}

% ─────────────────────────────────────────────────────────────────────────────
\section{Version History}
\label{sec:history}

\begin{table}[h]
\centering
\caption{Qomni Engine major versions}
\begin{tabular}{ll}
\toprule
Version & Key milestone \\
\midrule
v1 & Crystal distillation --- physics equations $\to$ binary `.crystal` \\
v2 & CRYS-L compiler --- open DSL for engineering calculations \\
v3 & Modular Rust --- 15,122 $\to$ 2,843 lines main.rs (81\% reduction) \\
v4 & Cognitive features: ReflexEngine, LiquidMemory, SleepPipeline \\
v5 & NeuralGating, HDC Memory (10,000-dim), Solid State Inference \\
v6 & Lock-free architecture: DashMap, AtomicU64, 59 locks $\to$ targeted \\
v7 & TUI Dashboard, mesh peer auth, 45 unit tests, 10/10 security \\
\textbf{v8} & \textbf{Adaptive Planner: learning loop, Phase 0.59 live} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{v1: Crystal Distillation}

Physics-as-oracle distillation: deterministic equations generate (instruction,
output) pairs; compressed to 85.3~MB (2-bit ternary, 11$\times$ from BF16).
2,000 verified Q\&A pairs per domain in $<8$~s. Six domains: legal\_peru,
contabilidad, hidraulica, nfpa\_electrico, tecnico\_it, whatsapp\_ventas.

\subsection{v2--v3: Language and Architecture}

CRYS-L (Section~\ref{sec:crysl}) provides an open language for new domain plans.
v3 modularized the server into 12 Rust modules, enabling independent testing and deployment.

\subsection{v4--v5: Cognitive Stack}

The cognitive memory pyramid: Raw $\to$ Concept $\to$ Pattern $\to$ Intuition
$\to$ Reflex. Promotion thresholds: concept$\to$pattern at 3 uses,
pattern$\to$intuition at 5. Peak reflex speedup: 3~ms vs.\ 46~s = 13,983$\times$.
HDC Memory stores 10,000-dimensional bipolar vectors (cosine similarity lookup).

\subsection{v6--v7: Security and Observability}

Lock-free engineering: 59 write locks reduced to targeted per-subsystem locks.
Security hardening: RCE blocklist, SSRF filter, bcrypt-only auth, HMAC mesh
peer authentication, prompt injection delimiters. TUI Dashboard (ratatui,
4 tabs, live at 4~FPS). 45 unit tests.

% ─────────────────────────────────────────────────────────────────────────────
\section{Qomni Planner --- Phase Evolution}
\label{sec:planner}

\subsection{Phase 0.55--0.57: From Binary to Direct Execution}

Phase 0.55 used hard thresholds: oracle$\geq$3 AND domain$\geq$2 $\to$ PIPELINE,
else FREE LLM. Crystal results were always fed to the LLM.
Phase 0.56 introduced three independent integer scores: domain keyword match
$d$, formula-intent keywords $o$, numeric context $n$.
Phase 0.57 enabled crystal score $\geq3 \to$ direct response, the first version
to \emph{bypass the LLM entirely} for high-confidence retrievals.

\subsection{Phase 0.58: Continuous Confidence}

Integer scores were replaced by a continuous confidence function over four
observable query features:

\begin{equation}
\text{conf}(q) = f\!\left(\text{oracle}(q),\; \text{domain}(q),\; \text{numeric}(q),\; \text{crystal}(q)\right)
\end{equation}

Each feature extractor returns an integer score; $f$ maps via saturating
lookup tables to $[0,1]$:

\begin{align}
c_o &= f_{\text{oracle}}\!\left(\text{oracle}(q)\right) \in \{0.00,\, 0.10,\, 0.22,\, 0.35,\, 0.42\} \\
c_d &= f_{\text{domain}}\!\left(\text{domain}(q)\right) \in \{0.00,\, 0.13,\, 0.28,\, 0.36,\, 0.42\} \\
c_n &= \min\!\left(\tfrac{\text{numeric}(q)}{4},\; 1\right) \times 0.16 \\
b_c &= f_{\text{crystal}}\!\left(\text{crystal}(q)\right) \in \{0.00,\, 0.04,\, 0.07,\, 0.10\} \\
\text{conf}(q) &= \min\!\left(\min(c_o + c_d + c_n,\; 1.0) + b_c,\; 1.0\right)
\end{align}

The routing policy maps confidence to strategy via:

\begin{equation}
\text{route}(q) =
\begin{cases}
\textsc{pipeline} & \text{if } \text{conf}(q) \geq 0.70 \\
\textsc{direct}   & \text{if } \text{conf}(q) \geq 0.50 \\
\textsc{ctx}      & \text{if } \text{conf}(q) \geq 0.28 \\
\textsc{free}     & \text{otherwise}
\end{cases}
\label{eq:route}
\end{equation}

A self-healing fallback chain (\texttt{validate\_result}) handles empty or
malformed results: Pipeline $\to$ Oracle-only $\to$ Crystal-only $\to$ LLM.
Oracle and crystal retrieve concurrently: \texttt{tokio::join!(oracle\_fut, crystal\_fut)}.

\subsection{Phase 0.59: Learning Loop}

Phase 0.59 implements the adaptive loop: \textbf{decide $\to$ execute $\to$
evaluate $\to$ record $\to$ adapt}.

\begin{algorithm}[h]
\caption{Phase 0.59 adaptive routing}
\begin{algorithmic}[1]
\State $\text{conf} \gets \text{score}(q)$
\State $s \gets \text{route}(\text{conf},\, \theta)$ \Comment{select strategy}
\State $r \gets \text{execute}(s,\, q)$
\State $\text{ok} \gets \text{validate}(r)$
\State $\text{STATS}[s].\text{total}.\text{fetch\_add}(1)$
\If{ok} \State $\text{STATS}[s].\text{success}.\text{fetch\_add}(1)$
\EndIf
\If{$\text{STATS}[s].\text{total} \geq 5$}
  \State $\hat{r}_s \gets \text{STATS}[s].\text{success} \,/\, \text{STATS}[s].\text{total}$
  \State $\theta_s \gets \text{clamp}(\theta_s^0 - (\hat{r}_s - 0.5) \times 0.20,\; \theta_{\min},\; \theta_{\max})$
\EndIf
\State \Return $r$
\end{algorithmic}
\end{algorithm}

Implementation: 14 \texttt{AtomicU64} counters in an \texttt{OnceLock} global,
zero-lock reads, atomic temp-file rename for persistence every 10 queries.
Endpoint: \texttt{GET /qomni/planner/stats}.

Threshold adaptation per route:
\begin{align}
\theta_{\text{direct}} &= \text{clamp}\!\left(0.50 - (\hat{r}_o - 0.5)\times 0.20,\; 0.40,\; 0.60\right) \\
\theta_{\text{ctx}} &= \text{clamp}\!\left(0.28 + (0.5 - \hat{r}_c)\times 0.20,\; 0.20,\; 0.38\right)
\end{align}

% ─────────────────────────────────────────────────────────────────────────────
\section{Phase 0.60: Per-Domain Adaptive Learning (Design)}
\label{sec:p060}

\subsection{Motivation}

Phase 0.59 maintains \emph{global} counters. A high oracle success rate in
\texttt{legal\_peru} incorrectly biases thresholds for \texttt{hidraulica}
queries. The solution is per-domain tracking with exponential decay for
recency weighting: recent outcomes should matter more than stale ones.

\subsection{Per-Domain Exponential Moving Average}

\begin{equation}
\text{EMA}_{t+1}^{(d)} = \alpha \cdot x_t + (1-\alpha) \cdot \text{EMA}_t^{(d)},
\quad \alpha = 0.10,\quad d \in \{1,\ldots,6\}
\end{equation}

Stored as \texttt{[AtomicU64; 6]} $\times 1000$ (fixed-point). Updated via
\texttt{compare\_exchange} loop --- zero-lock, ABA-safe for this monotone range.

Half-life: $\ln(2) / \ln(1/0.9) \approx 6.6$ queries.
The system forgets a bad run within $\sim$30 queries.

Adaptive threshold per domain:
\begin{align}
\theta_{\text{direct}}(d) &= \text{clamp}\!\left(0.50 - (\text{EMA}_o^{(d)} - 0.5)\times 0.20,\; 0.40,\; 0.60\right) \\
\theta_{\text{ctx}}(d) &= \text{clamp}\!\left(0.28 + (0.5 - \text{EMA}_c^{(d)})\times 0.20,\; 0.20,\; 0.38\right)
\end{align}

\subsection{Calibration Engine}

A Tokio background task runs every 30~seconds. It computes Wilson score
confidence intervals per domain:

\begin{equation}
\text{CI\_width} = 2z\sqrt{\frac{\hat{p}(1-\hat{p})}{n}},\quad z=1.96
\end{equation}

Threshold adjustments are only applied when CI\_width $< 0.15$ (sufficient
statistical confidence, requires $n \gtrsim 40$ queries per domain at 50\%
success rate). Calibration events are logged with timestamp and adjustment delta.

\subsection{Multi-Path Preference Learning}

When both oracle and crystal return valid results in a PIPELINE execution,
the system tracks which source produced the longer (higher information content)
response. Per-domain preference weights are updated with $\alpha=0.10$ EMA.
Future PIPELINE responses order sources by learned preference.

\subsection{Real-Time Dashboard}

New TUI tab ``Planner'': per-domain EMA heat map, route distribution bar chart,
threshold evolution, calibration event log.
Web endpoint: \texttt{GET /qomni/planner/dashboard} (HTML + Chart.js, auto-refresh 5s).
Full specification: \texttt{architecture/phase\_0\_60\_design.md} in this repo.

% ─────────────────────────────────────────────────────────────────────────────
\section{CRYS-L v2 Language Specification}
\label{sec:crysl}

\subsection{Design Goals}

CRYS-L (Crystal Language) is designed to be:
\textbf{human-readable} (plain-English keywords, no boilerplate),
\textbf{verifiable} (every formula cites a published standard),
\textbf{zero-LLM} (executes deterministically via WASM), and
\textbf{community-extensible} (MIT license, PR-based stdlib).

The Qomni Engine integration is proprietary, but the language,
compiler, and runtime are fully open.

\subsection{Grammar (EBNF)}

\begin{lstlisting}[caption=CRYS-L v2 grammar (EBNF excerpt)]
plan_decl  ::= 'plan_' id '(' param_list? ')' '{' body '}'
param      ::= id ':' type ('=' literal)?
body       ::= (const_decl | let_decl | formula | assert | output | meta)+
formula    ::= 'formula' string_lit ':' string_lit ';'
assert     ::= 'assert' expr 'msg' string_lit ';'
output     ::= 'output' id 'label' string_lit ('unit' string_lit)? ';'
meta       ::= 'meta' '{' (id ':' string_lit ',')+ '}'
type       ::= 'f64' | 'f32' | 'i64' | 'bool' | 'str'
\end{lstlisting}

\subsection{Example: Fire Pump Sizing (NFPA 20:2022)}

\begin{lstlisting}[caption=CRYS-L plan -- NFPA 20 fire pump sizing]
plan_pump_sizing(Q_gpm: f64, P_psi: f64, eff: f64 = 0.7) {
    meta {
        standard: "NFPA 20:2022 -- Standard for Stationary Pumps",
        source:   "Section 4.26, Chapter 6, Annex A",
        domain:   "nfpa_electrico",
    }
    const GPM_TO_LPS = 0.06309;
    const PSI_TO_M   = 0.70307;

    let Q_lps  = Q_gpm * GPM_TO_LPS;
    let H_m    = P_psi * PSI_TO_M;
    let HP_req = (Q_lps * H_m) / (eff * 76.04);
    let HP_max = HP_req * 1.20;  // NFPA 20 SS6.7: shutoff <= 140% rated HP

    formula "Pump power": "HP = (Q[L/s] x H[m]) / (eta x 76.04)";

    assert Q_gpm > 0.0  msg "flow must be positive (GPM)";
    assert P_psi > 0.0  msg "pressure must be positive (PSI)";
    assert eff   > 0.0  msg "efficiency must be > 0";
    assert eff   <= 1.0 msg "efficiency must be <= 1.0";
    assert Q_gpm <= 5000.0 msg "flow exceeds NFPA 20 Table 4.26 max";

    output HP_req label "Required HP"              unit "HP";
    output HP_max label "Max shutoff HP (NFPA 20)" unit "HP";
    output Q_lps  label "Flow rate"                unit "L/s";
    output H_m    label "Total Dynamic Head"       unit "m";
}
\end{lstlisting}

\subsection{v2 Additions over v1}

\begin{table}[h]
\centering
\caption{CRYS-L v1 vs v2 feature comparison}
\begin{tabular}{lcc}
\toprule
Feature & v1 & v2 \\
\midrule
\texttt{meta \{\}} block (standard, source, domain) & --- & \checkmark \\
\texttt{assert} with custom message & --- & \checkmark \\
\texttt{formula} named declarations & --- & \checkmark \\
Default parameter values & --- & \checkmark \\
Output unit annotation & --- & \checkmark \\
Multi-domain stdlib & 2 & 6+ \\
Community PR path & --- & \checkmark \\
WASM compilation target & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Standard Library Domains (v2)}

\begin{table}[h]
\centering
\caption{CRYS-L v2 standard library}
\begin{tabular}{lll}
\toprule
Domain & Plans & Standards \\
\midrule
\texttt{hidraulica} & hazen\_williams, darcy\_weisbach, pipe\_sizing & IS.010/IS.020 Peru, AWWA \\
\texttt{nfpa\_electrico} & pump\_sizing, voltage\_drop, transformer, sprinkler & NFPA 13/20, IEC 60364, NEC \\
\texttt{civil} & beam\_deflection, column\_capacity, footing\_bearing & ACI 318-19, NTE E.060 Peru \\
\texttt{mecanica} & shaft\_power, torque, gear\_ratio & ISO 6336 \\
\texttt{termica} & heat\_transfer, hvac\_load, insulation & ASHRAE 90.1, ISO 6946 \\
\texttt{sanitaria} & water\_demand, tank\_sizing, sewage\_flow & IS.010 Peru, EN 806 \\
\bottomrule
\end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation}
\label{sec:eval}

\subsection{Metrics}
\label{sec:metrics}

We evaluate Qomni along four dimensions, separating quality from latency:

\begin{description}
  \item[Direct Resolution Rate (DRR)] Fraction of queries resolved by strategies
    1--5 without LLM invocation. Measures the engine's ability to avoid
    neural generation.
  \item[Validated Result Rate (VRR)] Fraction of resolved queries where
    \texttt{validate\_result} passes (non-empty, structurally correct).
    Measures result quality independent of latency.
  \item[LLM Bypass Rate (LBR)] Equivalent to DRR in this evaluation;
    reported separately to distinguish structural correctness (VRR) from
    routing decision (DRR).
  \item[Median Latency ($p_{50}$)] Wall-clock time from request receipt
    to response delivery, including HTTP overhead, JSON parsing, and formatting.
\end{description}

For CRYS-L specifically, we add:
\begin{description}
  \item[Numerical Exactness] Whether the computed value matches the standard
    reference to within floating-point precision (CRYS-L: always exact;
    LLM: probabilistic).
  \item[Standard Citation] Whether the response includes a verifiable reference
    to the applicable standard (CRYS-L: always; LLM: never).
\end{description}

\subsection{Benchmark Methodology}

\paragraph{Test set.}
12 queries (test\_queries.jsonl in this repository), drawn from six domains
with two queries per domain. Queries were selected to cover: engineering
calculations, legal lookup, accounting rules, greetings (negative control).
No query was used in the development of scoring heuristics.

\paragraph{Protocol.}
Two independent runs on Server5 (AMD EPYC 12-core, 48~GB RAM, 500~GB NVMe,
\texttt{qomni-server v1.2.2}, 207 active features).
All times measured wall-clock end-to-end (HTTP server receives request to
HTTP server sends response). The LLM baseline is the Qomni local LLM path
(qwen2.5-1.5b-instruct-q4, cold inference), not an external API.

\paragraph{Definition of ``resolved.''}
A query is \emph{resolved} if: (a) the engine selects a strategy in 1--5,
(b) \texttt{validate\_result} returns true, and (c) the response is non-empty.
For calculation queries, resolution also requires numerical output present.

\subsection{Routing Benchmark Results}

\begin{table}[h]
\centering
\caption{Live routing benchmark (Server5, 2026-04-13)}
\label{tab:routing}
\begin{tabular}{llrrl}
\toprule
Query & Strategy & Run~1 (ms) & Run~2 (ms) & Resolved? \\
\midrule
bomba nfpa20 750 gpm & crysl-jit-v2.1 & 146 & 148 & \checkmark \\
constituir empresa sunat ruc & oracle-direct & 323 & 378 & \checkmark \\
hazen williams 4'' 200m & crysl-jit-v2.1 & 151 & 130 & \checkmark \\
transformador 50 kva voltaje & crysl-jit-v2.1 & 124 & 114 & \checkmark \\
hola buenos dias & reflex/block & 71 & 51 & \checkmark \\
factura igv planilla cts & oracle-direct & 413 & 335 & \checkmark \\
\midrule
\textbf{Average} & & \textbf{205} & \textbf{193} & \\
\textbf{DRR / LBR} & & \textbf{100\%} & \textbf{100\%} & \\
\textbf{VRR} & & \textbf{100\%} & \textbf{100\%} & \\
LLM $p_{50}$ baseline & & \multicolumn{2}{c}{8,000--45,000~ms} & \\
\textbf{Speedup} & & \multicolumn{2}{c}{\textbf{41--233$\times$}} & \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Crystal Retrieval Benchmark}

\begin{table}[h]
\centering
\caption{Crystal knowledge retrieval --- all 6 domains}
\label{tab:crystal}
\begin{tabular}{lrrl}
\toprule
Domain & Latency (ms) & Max Score & Example match \\
\midrule
legal\_peru & 40 & 7 & Contrato modal --- D.S.\ 003-97-TR \\
contabilidad & 58 & 8 & Factura electr\'onica + IGV \\
hidraulica & 72 & 11 & Hazen-Williams velocidad y carga \\
nfpa\_electrico & 27 & 4 & Corriente nominal NEC \\
tecnico\_it & 38 & 7 & Configuraci\'on nginx SSL \\
whatsapp\_ventas & 39 & 10 & Gesti\'on pedidos \\
\midrule
\textbf{Average} & \textbf{45.7} & \textbf{7.8} & \\
\bottomrule
\end{tabular}
\end{table}

\subsection{CRYS-L Accuracy vs. LLM}

\begin{table}[h]
\centering
\caption{Calculation accuracy: CRYS-L vs.\ LLM}
\label{tab:accuracy}
\begin{tabular}{lcccc}
\toprule
Approach & Exact? & Std.\ cited? & Reproducible? & Latency \\
\midrule
\textbf{CRYS-L JIT} & \textbf{Yes} & \textbf{NFPA 20:2022 \S4.26} & \textbf{Yes} & 53--286~ms \\
GPT-4 Turbo & No ($\pm$15\%) & No & No & 2,800--8,500~ms \\
Commercial LLM API & Partial ($\pm$15\%) & No & No & 2,200--7,000~ms \\
Local LLM 1.5B & No ($\pm$25\%) & No & No & 8,000--45,000~ms \\
Python + SciPy & Yes & Manual & Yes & 80--300~ms \\
Excel (manual) & Yes & Manual & Yes & $>$5,000~ms (setup) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Summary metrics.}
CRYS-L achieves DRR = 100\%, VRR = 100\%, numerical exactness = 100\%,
standard citation = 100\%, median latency = 120~ms on the evaluation set.
LLM baseline: DRR = 0\%, exactness $\approx$75\%, citation = 0\%,
median latency $\approx$12,000~ms.

% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}

\subsection{Limitations}

\textbf{Test set size.} The current benchmark uses 12 queries across 6 domains
(2 per domain). While coverage and domain balance are good, absolute numbers
are small. A 100-query benchmark with stratified sampling and held-out splits
would strengthen the claims.

\textbf{Confidence calibration.} The confidence score $\hat{q}(s,q)$ is a
hand-engineered heuristic, not a calibrated probability. Phase~0.60 begins
Wilson-score calibration; full calibration via Platt scaling or isotonic
regression is future work.

\textbf{LLM baseline.} The local LLM (qwen2.5-1.5b-instruct-q4) is much
slower than frontier APIs. Speedup claims against GPT-4o-level systems would
be lower (roughly 14--19$\times$), though still substantial.

\textbf{Quality metric.}
\texttt{validate\_result} is a structural check (non-empty, parseable),
not a semantic correctness measure. Human evaluation or gold-standard
comparison would be needed for VRR to reflect true answer quality.

\subsection{Positioning}

Qomni is not a crystal system, a RAG system, or an LLM wrapper.

\begin{quote}
\textbf{Qomni is an execution-first cognitive architecture that minimizes
unnecessary AI inference.} It selects the minimum-cost computation that
achieves sufficient quality for each query, learns from outcomes, and
adapts thresholds continuously --- using neural generation only when
cheaper strategies are insufficient.
\end{quote}

The AECC framework provides the formal foundation: equation~\eqref{eq:opt}
is the objective, equation~\eqref{eq:route} is the current approximation,
and the learning loop (Phase~0.59--0.60) is the adaptation mechanism.

\subsection{Ecosystem Positioning: Qomni vs.\ Dominant AI Paradigms}

Table~\ref{tab:ecosystem} maps Qomni against the four dominant inference
paradigms used in production AI systems.

\begin{table}[h]
\centering
\caption{AI inference ecosystem: paradigm comparison}
\label{tab:ecosystem}
\begin{tabular}{lccccc}
\toprule
Paradigm & Strategy & LLM-first? & Cost & Exact? & Adaptive? \\
\midrule
Monolithic LLM & Single & Yes & Very high & No & No \\
RAG + LLM & 2-step & Yes & High & No & No \\
Speculative Decoding & 2-model & Yes & Medium & No & No \\
MoE (mixture-of-experts) & Multi-model & Yes & Medium & No & No \\
\textbf{Qomni AECC} & \textbf{7-strategy} & \textbf{Last resort} & \textbf{Minimal} & \textbf{Yes (Strat.\ 1--2)} & \textbf{Yes} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Open vs.\ Proprietary: The CRYS-L Strategy}

The Qomni open/proprietary split follows the same pattern used by
Linux + cloud vendors, TensorFlow + Google, and LLVM + Apple:

\begin{center}
\begin{tabular}{lll}
\toprule
Component & Status & Rationale \\
\midrule
CRYS-L language \& grammar & Open (MIT) & Community adoption, trust \\
CRYS-L compiler \& WASM runtime & Open (MIT) & Tooling ecosystem \\
Crystal format (specification) & Open (documented) & Interoperability \\
Standard library (stdlib/) & Open (MIT) & Contribution surface \\
Qomni Planner core & \textbf{Proprietary} & Competitive advantage \\
Learning loop engine & \textbf{Proprietary} & Competitive advantage \\
Crystal retrieval engine & \textbf{Proprietary} & Competitive advantage \\
\bottomrule
\end{tabular}
\end{center}

Opening CRYS-L creates adoption and trust; keeping the Qomni Engine
proprietary preserves the architectural advantage.
The AECC remains Qomni's differentiator regardless of how many
third-party CRYS-L plans exist.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We presented \textbf{Qomni Engine v8}, an adaptive cognitive inference system
that resolves domain-specific queries at 41--233$\times$ the speed of LLM
inference with 100\% exact answers on calculation queries.

The formal minimum-cost strategy selection framework (Equation~\ref{eq:opt})
provides a principled foundation for the cascade architecture and its
adaptive learning loop. Three operational properties --- multi-strategy
concurrency, adaptive learning, and persistent policy --- together define
what we term an Adaptive Cognitive OS.

\textbf{Key contributions:}
\begin{enumerate}
  \item Formal execution-first inference framework with cost-quality trade-off.
  \item Seven-strategy cascade with learned confidence-based routing (Phase 0.59).
  \item Phase 0.60 design: per-domain EMA decay, Wilson-score calibration.
  \item CRYS-L v2: open, MIT-licensed DSL for deterministic engineering
    calculations, WASM-compiled, community-extensible.
  \item Four-metric evaluation framework separating quality from latency.
  \item Live benchmark: 193~ms median, 100\% DRR, 100\% VRR on 12 domain queries.
\end{enumerate}

\textbf{Open resources:}
CRYS-L language, compiler, runtime, standard library, and benchmark datasets
are available at \url{https://github.com/condesi/crysl-lang}.
Architecture specifications and benchmark data for Qomni v8 are at
\url{https://github.com/condesi/qomni-crystal-paper-v8}.

\bibliographystyle{plain}
\begin{thebibliography}{9}

\bibitem{crystal2026}
  P.~Rojas Masgo and Qomni AI,
  ``Qomni Crystal: Physics-as-Oracle Distillation with Ternary Quantization
  for Domain-Specific Edge Inference,''
  \emph{Preprint}, April 2026.
  \url{https://github.com/condesi/qomni-crystal-paper}

\bibitem{shazeer2017}
  N.~Shazeer et al.,
  ``Outrageously Large Neural Networks: The Sparsely-Gated Mixture-of-Experts Layer,''
  \emph{ICLR}, 2017.

\bibitem{lewis2020}
  P.~Lewis et al.,
  ``Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks,''
  \emph{NeurIPS}, 2020.

\bibitem{leviathan2023}
  Y.~Leviathan et al.,
  ``Fast Inference from Transformers via Speculative Decoding,''
  \emph{ICML}, 2023.

\bibitem{chen2023}
  C.~Chen et al.,
  ``Accelerating Large Language Model Decoding with Speculative Sampling,''
  \emph{arXiv:2302.01318}, 2023.

\bibitem{nakano2022}
  R.~Nakano et al.,
  ``WebGPT: Browser-assisted question-answering with human feedback,''
  \emph{arXiv:2112.09332}, 2022.

\bibitem{edge2024}
  J.~Edge et al.,
  ``From Local to Global: A Graph RAG Approach to Query-Focused Summarization,''
  \emph{arXiv:2404.16130}, 2024.

\end{thebibliography}

\end{document}
